---
title: "Urban data lab - day 2 - load data from WFS - charging stations HH"
author: "Sven Lautenbach"
date: "10/18/2020"
output: 
  html_document: 
    toc: yes
    highlight: espresso
    number_sections: yes
    theme: sandstone
---

```{r setup, include=FALSE}
# you need to adapt this path to your local machine
# Windows users be careful: use the slash '/' not the backslash '\'
# as the path separator
ws <-"/home/slautenb/Documents/lehre/HD/ws_2020_21/heikaLab/R4UrbanDataLab_2020/" 
require(knitr)
opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = ws)
# handling of spatial data
require(sf)
require(tmap) # map creation
require(ows4R) # acces OGC webservices

require(httr)
require(tidyverse) # 
require(ggplot2)
```

# Preparation

Ensure that local folder structure exists

```{r}
dataFolder = paste0(ws, "/data/")
if(!dir.exists(dataFolder))
  dir.create(dataFolder)
```

# Getting the data from a WFS

```{r}
wfsUrl <- "https://geodienste.hamburg.de/HH_WFS_EMobility"
```

Connect to the service

```{r}
hh_client <- WFSClient$new(wfsUrl, 
                            serviceVersion = "2.0.0")
```

What are the capabilities of the service?

```{r}
hh_client$getCapabilities()
```


Which features are present?
```{r}
hh_client$getFeatureTypes(pretty = TRUE)
```
Get the features from app:stromnetz_emobility
```{r}
wfsUrlParsed <- parse_url(wfsUrl)
wfsUrlParsed$query <- list(service = "wfs",
                  version = "2.0.0",
                  request = "GetFeature",
                  typename = "app:stromnetz_emobility",
                  srsname='urn:x-ogc:def:crs:EPSG:25832',
                  bbox = "552378.61, 5921782.142, 582209.533, 5956674.616")
request <- build_url(wfsUrlParsed)

chargingStations <- read_sf(request) 
```

See what we have got:

```{r}
names(chargingStations)
```

The CRS did not make it through the request so we need to manually assign the CRS. The meta-data of the service indicate that it is EPSG:25832 (UTM zone 32N).

```{r}
st_crs(chargingStations) <- 25832
```


```{r}
tmap_mode("view")
tm_shape(chargingStations) + tm_dots(col = "anzahl_ladepunkte", size=0.5, alpha=.5) + tm_basemap(server= "OpenStreetMap.DE")
```
We could even get information on the current status of the charging station.

```{r}
tmap_mode("view")
tm_shape(chargingStations) + tm_dots(col = "ladesaeule_status", size=0.5, alpha=.5) + tm_basemap(server= "OpenStreetMap.DE")
```

## Save data to geopackage

We will not be using the live data but just the locations in the following. Therefore, it is fine to save the data - which might save us some time later on.

```{r}
st_write(chargingStations, dsn= paste0(dataFolder, "HH.gpkg"), layer="chargingStations", delete_layer=TRUE)
```

