---
title: "Urban data lab - day 2 - explorative analysis - charging stations HH"
author: "Sven Lautenbach"
date: "10/18/2020"
output: 
  html_document: 
    toc: yes
    highlight: espresso
    number_sections: yes
    theme: sandstone
---

```{r setup, include=FALSE}
# you need to adapt this path to your local machine
# Windows users be careful: use the slash '/' not the backslash '\'
# as the path separator
ws <-"/home/slautenb/Documents/lehre/HD/ws_2020_21/heikaLab/R4UrbanDataLab_2020/" 
require(knitr)
opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = ws)
# handling of spatial data
require(sf)
require(tmap) # map creation
require(tmaptools)
require(raster) # handling of raster data, map algebra etc.
require(KernSmooth) # kernel density estimation
require(dbscan) # dbscan clustering
require(OpenStreetMap) # for offline background maps. Depends on rJava that might be a bit tricky to install. Functionality is not essential, you might just outcomment the lines that require OpenStreetMap-package functionality
#require(tidyverse) # data preprocessing
#require(ggplot2) # plotting functionality
```

# Preparation

```{r}
dataFolder <- paste0(ws, "data/")
```


Load data from geopackage


```{r}
charging_stations <- st_read(dsn= paste0(dataFolder, "HH.gpkg"), layer="chargingStations")
```

# Explorative analysis

## Kernel density map

We need to creat/define an empty raster that will store the results of the kernel density estimation

Get bounding box of the data

```{r}
(bboxHH <- st_bbox(charging_stations))
```


```{r}
gridsize=c(500,500)
range_x=list(c(bboxHH[1], bboxHH[3]),c(bboxHH[2], bboxHH[4]))
#552378.61, 5921782.142, 582209.533, 5956674.616
kdeMat <-  bkde2D(st_coordinates(charging_stations), bandwidth=c(500,500), gridsize=gridsize, range.x=range_x)
```

Since *bkde2D* returns a matrix we need to convert it to a raster:

```{r}
kdeRas <- raster(list(x=kdeMat$x1,y=kdeMat$x2,z=kdeMat$fhat), crs=CRS('+init=EPSG:25832') )
```

Scale the raster so that units are more meaningful and create countour lines that we will use in plotting

```{r}
kdeRas_scaled <- kdeRas*xres(kdeRas)*yres(kdeRas)* nrow(charging_stations)
cellStats(kdeRas_scaled$layer, stat=sum )
kdeRas_scaledContour <- rasterToContour(kdeRas_scaled)  %>%   st_as_sf()
```


```{r}
tmap_mode("view")
tm_shape(kdeRas_scaled) +
  tm_raster(col="layer", alpha = .5, palette ="-plasma", title = "Kernel density estimate", style = "kmeans" ) + 
  tm_shape(charging_stations) + 
  tm_dots(alpha=.5) + 
  tm_shape(kdeRas_scaledContour) + 
  tm_lines() + 
  tm_text(text="level", col="white", shadow=TRUE, along.lines=TRUE)
```

## How far are the stations away from each other?

We use a for loop to calculate the distance to the clostet charging station in the data set. Therefore, we loop over all points and calculate the distance between the point i and all points including itself. This rather brute force approache returns the whole distance matrix which when can be used to calculate differences.


```{r}
n <- nrow(charging_stations)
distMat <- matrix(data=0, nrow=n, ncol=n)
for(i in seq_len(n))
{
  allDists <- st_distance(charging_stations, charging_stations[i,])
  distMat[i,] <- allDists
  
}
```

In many cases several charging stations are at the same point, rendering a distance of zero.

```{r}
getSecondSmallest <- function(x, pos=2)
{
   sort(x)[pos]
}
dist2Shortest <- apply(distMat, MARGI=1, FUN= getSecondSmallest, pos=2)
summary(dist2Shortest)
```

We can force the analysis to deliver us the shortest distance to a neighbor not at the same location. Therefore, we:
  * sort values
  * get the indices of all values larger than zero
  * get the first element of the indices that is larger then zero

```{r}
getSecondSmallestNotAtSame <- function(x, pos=1)
{
   sortedValues <- sort(x)
   idx <- which(sortedValues > 0)
   return(sortedValues[idx][pos])
}
dist2ShortestNotAtSame <- apply(distMat, MARGI=1, FUN= getSecondSmallestNotAtSame)
summary(dist2ShortestNotAtSame)
```

Or we could get the average distance to the closest ten charging stations


```{r}
getMeanDistToClostest <- function(x, n=10)
{
   sortedValues <- sort(x)
   return(mean(sortedValues[1:n+1]))
}
distMean10 <- apply(distMat, MARGI=1, FUN= getMeanDistToClostest)
summary(distMean10)
```

If we want to show distances in the map we can simply add the vectors to the simple features object and use tmap afterwards.

```{r}
charging_stations['meanDist10'] <- distMean10
```

```{r}
osm_hh <- read_osm(charging_stations, ext=1.1)
```


```{r}
tmap_mode("plot")
tm_shape(osm_hh) + tm_rgb() + 
  tm_shape(charging_stations) + 
  tm_dots(size="meanDist10", legend.size.show = TRUE, title.size = "Mean distance to closest 10 stations" ) + 
  tm_basemap(server= "OpenStreetMap.DE") + 
  tm_layout(legend.outside=TRUE) + 
  tm_scale_bar()
```

We see that distances are (not unexpectedly increaing from the city center).

A convenient and fast imlementations is *kNNdist* from the *dbscan* package:

```{r}
nn4Mat <- kNNdist(st_coordinates(charging_stations), k=4, all=TRUE)
head(nn4Mat)
```

It returns the distance matrix for the *k* nearest neighbors.


## Identify spatial clusters using dbscan

DBSCAN estimates the density around each data point by counting the number of points in a user-specified eps-neighborhood and applies a used-specified minPts thresholds to identify core, border and noise points. In a second step, core points are joined into a cluster if they are density-reachable (i.e., there is a chain of core points where one falls inside the eps-neighborhood of the next). Finally, border points are assigned to clusters. The algorithm only needs parameters eps and minPts.


```{r}
db_hh <- dbscan(st_coordinates(charging_stations), eps = 650, minPts = 7)
print(db_hh)
```

```{r}
charging_stations['dbCluster'] <- db_hh$cluster
nClust <- length(unique(db_hh$cluster))
```

```{r}
tmap_mode("view")
tm_shape(charging_stations) + 
  tm_dots(col = "dbCluster", size=0.25, alpha=.5, palette="Accent", n=nClust) +
  tm_basemap(server= "OpenStreetMap.DE") + 
  tm_scale_bar()
```

# Exercise

Perform the same anlysis with the dataset chargingStations_berlin.shp. You can load the shapefile by simply passing the folder name and the name of the shapefile (including the .shp extension) to st_read.
The data are in ETRS89 / UTM zone 33N (use a web search to get the EPSG number)